% main.tex -- Article 2 (IEEEtran). Skeleton; full assembly in Phase 7.1.
% Reuses the Article 1 toolchain (CLAUDE.md S9). Compile:
%   eval "$(/usr/libexec/path_helper)"; latexmk -pdf -synctex=1 -interaction=nonstopmode main.tex
% NOTE: \usepackage{url} is REQUIRED -- the BibTeX .bbl wraps Scopus URLs (with % and &)
% in \url{}; without the url package the bbl's \providecommand{\url}[1]{#1} treats % as a
% comment and swallows the bibliography (runaway argument). Do not remove it.
\documentclass[lettersize,journal]{IEEEtran}
\usepackage[table]{xcolor}
\usepackage{array,amsmath,amssymb}
\usepackage{url}
\usepackage{pgf}        % figures are matplotlib .pgf, composed natively (Phase 7.3)
\def\mathdefault#1{#1}  % matplotlib pgf math tick labels (e.g. 10^2) without mpl's preamble
\usepackage{eurosym}    % \euro in Method/Results cost reporting
% Float tuning (7.4): 16 floats over ~9 text pages; defaults defer the big table* queue past
% the references. Allow fuller float pages/tops so tables interleave with their sections.
\setcounter{topnumber}{4}\setcounter{dbltopnumber}{3}\setcounter{totalnumber}{8}
\renewcommand{\topfraction}{0.95}\renewcommand{\dbltopfraction}{0.95}
\renewcommand{\textfraction}{0.05}
\renewcommand{\floatpagefraction}{0.75}\renewcommand{\dblfloatpagefraction}{0.75}

\begin{document}
\title{A Configuration Framework for LLM-Based Grading in Computer Science}

\author{Pedro~S.~Romano$^{\dagger}$, Daniel~F.~S.~Fernandes$^{*}$, and Pedro~J.~A.~Sebastião$^{*}$\\
\small $^{*}$Instituto de Telecomunicações, Lisbon 1049-001, Portugal\\
\small $^{\dagger}$Iscte -- Instituto Universitário de Lisboa, Lisbon 1649-026, Portugal}

\markboth{IEEE Transactions on Learning Technologies, Vol. XX, No. X, Month Year}%
         {Romano \MakeLowercase{\textit{et al.}}: A Configuration Framework for LLM-Based Grading in Computer Science}

\maketitle

% Register: CLAUDE.md S9.2 (no em/en dashes as prose punctuation). RQ1 fence: no
% "significant across all datasets"; the GPT-5.1 anchor never appears in the abstract.
\begin{abstract}
Instructors adopting graders based on large language models (LLMs) face a configuration problem:
should the model reason before scoring, what guidance should it receive, should an exam be graded in
one call or question by question, and which model is worth its cost? We study this design space
systematically for computer science assessment, crossing extended reasoning, evaluation guidance,
grading scope, rubric decomposition, model family, and conversation state under one protocol, on four
datasets covering short answers and code. Reasoning is isolated within each of three open-weight
model families, and every measurement is repeated, so its effect on cost and on run-to-run
consistency is measured alongside quality. Reasoning raised agreement with teacher grades in only
seven of twenty model-dataset combinations, at 10 to 925 times the output tokens and with reduced
consistency in nearly every cell. Evaluation guidance, tested within one model, helped
wherever the human reference was trustworthy; on grades from a real Portuguese deployment, an
unvalidated reference masked a real rubric benefit until evaluation was restricted to teacher-verified
grades. Transfer of public-dataset
results to that deployment was partial: the top configuration carried over, the mid-ranking did not.
We synthesise these comparisons into a decision guide mapping task characteristics to a recommended
configuration and its cost, governed by one rule: validate locally against a reference you trust. The
harness, prompts, and configurations are released.
\end{abstract}

\begin{IEEEkeywords}
Automated grading, large language models, learning analytics, computer science education,
short-answer grading, code assessment, configuration framework.
\end{IEEEkeywords}

% --- Sections (Phase 7.1) ---
\input{sections/introduction}
\input{related/sota_draft}            % Related Work -- drafted in Phase 1.6
\input{tables/published_baselines}    % Phase 1.5 context table (referenced from Related Work)
\input{sections/method}
\input{sections/results}
\input{sections/discussion}
\input{sections/threats}
\input{sections/conclusion}

\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
