%% sections/discussion.tex -- Phase 7.1, rewritten 2026-06-16. Register: CLAUDE.md S9.2 (no em/en
%% dashes as prose punctuation; one idea per sentence; numbers with meaning; paragraph openers carry
%% the point; refer back to Results, do not re-list every number). The framework artifact is presented
%% in its own section (sections/framework.tex, sec:framework); this section interprets the cross-cutting
%% findings and leads into it. Avoid the word "gold" and the "nearly published" framing.
%% (SemEval seen/unseen splits CUT 2026-06-15: do not reintroduce the unseen-domain observation.)

\section{Discussion}
\label{sec:discussion}

Read together, the six questions return one message: in LLM grading almost nothing is unconditional.
Whether a configuration choice helps depends on the task at hand, the model's behaviour, the reliability
of the grades used as reference, and which quality is being optimised, whether agreement, consistency, or
cost. The paragraphs below draw the findings out along these lines and then take up what they imply for
choosing a configuration.

Reasoning was the clearest conditional effect. Extended thinking improved agreement with teacher grades
in only a minority of model and dataset cells, and it always came with a high price. It multiplied output tokens by
between 10 and 925 times, with the premium itself depending on the domain, and in most cells it lowered
run-to-run consistency, so the same answer scored itself less alike across repetitions. That consistency
loss is only partly mechanical: longer outputs give a non-deterministic backend more room to diverge, but
the link between an answer's length and its grade variability was moderate and uneven, so part of the
loss behaves like a property of reasoning itself rather than of length alone. Even the largest agreement
gain in the study was modest, about a quarter of a QWK point, while most cells gained little or nothing
and a few lost ground. Such a small return rarely justified a cost this large. The one dependable
exception was a grader whose baseline had collapsed to scoring much of the code at exactly zero; there,
reasoning restored the discrimination the baseline had lost. This is why the advice about
reasoning is keyed to an observable behaviour, whether the baseline collapses, rather than to a model's
name, a choice that also lets it outlast the next model release.

Evaluation guidance was the cheapest reliable improvement we found, raising agreement by 0.12 to 0.24 QWK
across the public datasets. The same benefit held on the teacher-reviewed subset of the deployment data,
where the rubric helped the code grader about as much as on the public code. Providing it is the one
configuration change in the study that reliably improved agreement.

How the grade was requested mattered far less. Grading a whole exam in one pass rather than question by
question, and scoring a rubric criterion by criterion rather than holistically, barely changed agreement
with the final grade, so these choices can be settled on cost and simplicity, with the cheaper shape the
better default. A whole exam here is about four questions, the size of these exams; we did not test
larger sets, so whole-exam grading should be read as a block of roughly that size rather than an
arbitrarily long one. The way the grading session is run had a small effect of a different kind. Grading
every answer inside one conversation that accumulated across the exam biased grades slightly downward
relative to giving each answer a clean conversation, and the order in which answers were presented added
variation with no consistent direction. Clean, question-by-question grading was thus both the least
biased and the most reproducible regime, which recommends it on fairness grounds.

Among the LLM models and configurations we tested, none graded best everywhere. On short answers the configurations clustered
within about 0.07 QWK of one another, so the choice is nearly free and is best made on cost; on code they
spread far more widely, about 0.26 QWK, and there the configuration is worth optimising. Across both
tasks the open models with reasoning off were the efficient choice, GLM-5.1 most of all, while the closed
frontier model never justified its much higher cost. The ranking learned on the public datasets carried
over at the top: the configuration that ranked first on public data also ranked first in deployment.
Below the top, the verified deployment samples were too small to tell the close-ranked configurations
apart, so their relative order remained statistically indeterminate. The natural choice is therefore the
configuration that ranks highest on public data: it is both the strongest grader there and the
configuration whose lead carried over to deployment.

Taken together, these results argue for a structured set of conditional choices rather than a single
recommended grader. The deciding factors recurred throughout the study: the task, the model's observable
behaviour, the reliability of the reference, and the quality being optimised. Two of the three quantities
that matter, consistency and cost, are routinely left unreported in other studies, yet both moved against
reasoning here, so reporting agreement alone would have hidden half of what a practitioner needs to
decide. Because the
choices turn on observable behaviour rather than on model identity, they also do not expire with each new
release. We therefore express the contribution not as a leaderboard but as a decision guide: a small set
of task-keyed priors, each carrying its own trade-off. The next section presents it.
