%% sections/introduction.tex -- Phase 7.1. Register: CLAUDE.md S9.2 (no em/en dashes as prose
%% punctuation; one idea per sentence; numbers with meaning; paragraph openers carry the argument).
%% RQ1 abstract/intro fence: never "significant across all datasets"; enumerate cells instead.
%% QWK is deliberately NOT used in this section: its first use, with the mandatory calibration
%% sentence, is in Method (PHASES 7.1).

\section{Introduction}
\label{sec:intro}

Large language models (LLMs) are moving from demonstrations to deployed graders in computer science
(CS) courses. The promise for learning analytics is well documented: rubric-conditioned LLMs approach
human raters on well-specified short-answer tasks, and they return criterion-level feedback at a scale
no teaching team can match~\cite{sessler_canai_2025,henkel_roars_2025}. The risk is equally well
documented. Grading errors distort how students revise their work~\cite{li_am_2023}, repeated scoring
of the same answer is unstable~\cite{pack_large_2024,moazzez_assessing_2024}, and agreement with
teachers varies sharply between models and prompts~\cite{tate_can_2024}. An instructor who adopts an
LLM grader is therefore not making one decision but several: whether the model should reason at length
before scoring, what evaluation guidance to place in the context, whether to grade a whole exam in one
call or question by question, whether to score a rubric criterion by criterion or holistically, and
which model family to run or pay for.

The literature answers these questions only in fragments. Reasoning is the clearest example. Extended
thinking improves general LLM-as-a-judge accuracy~\cite{jayarao_explicit_2025}, but its effect is task-
and architecture-dependent~\cite{larionov_deepseek_2025}, long reasoning traces can degrade
calibration-sensitive tasks~\cite{sui_stop_2025}, and no peer-reviewed study isolates its effect on
rubric-bounded educational grading. Model comparisons exist for programming and course
deployments~\cite{jukiewicz_comparison_2026,policar_automated_2025}, but they rank systems under one
fixed prompt configuration, so they cannot say how the ranking moves when the configuration does.
Guidance, scope, decomposition, and conversation state have each been touched as point solutions
(Section~\ref{sec:related}), never crossed under one protocol. What is missing is the practical
synthesis: for a given CS assessment task, which configuration should you use, and what does it cost?

This paper studies the configuration space systematically and delivers the synthesis as a decision
guide. We cross six design dimensions: extended reasoning (on or off), evaluation guidance (none
versus a rubric or reference answer), grading scope (whole exam versus question by question), rubric
decomposition (holistic versus criterion by criterion), model family, and conversation state (a clean
conversation per answer versus shared history). Three open-weight model families (Qwen3.5,
DeepSeek-V4-Flash, GLM-5.1) expose a within-family reasoning toggle, so the reasoning effect is
isolated from model identity, a confound in prior comparisons; a closed frontier model (GPT-5.1)
serves as a paid reference point. The study runs on four datasets in two domains: the Mohler corpus
and SemEval-2013 Task~7 for CS short answers, and the Rubric Is All You Need corpus plus PT-CS, an
export of a real Portuguese university grading deployment, for code. Every cell is repeated five times
(three for the paid model), so consistency is measured rather than assumed, and every call's token
cost is recorded.

Six research questions structure the study:
\begin{itemize}
  \item \textbf{RQ1.} How does extended reasoning affect grading quality, consistency, and cost, and
        does the effect differ between code and short answers?
  \item \textbf{RQ2.} How much does evaluation guidance (none versus a rubric or reference answer)
        change quality and consistency?
  \item \textbf{RQ3.} How do grading scope (whole exam versus question by question) and rubric
        decomposition (holistic versus criterion by criterion) affect agreement with the final grade?
  \item \textbf{RQ4.} How do current-generation models compare as graders, and what is the
        cost-quality trade-off between local open-weight and paid frontier models?
  \item \textbf{RQ5.} Does conversation state (clean versus shared history) affect consistency and
        fairness?
  \item \textbf{RQ6.} Does the configuration guidance found on public datasets transfer to a real
        Portuguese deployment context?
\end{itemize}

The findings motivate a framework rather than a leaderboard. Reasoning turned out to be a trade, not
an upgrade: it raised agreement with teacher grades in seven of the twenty model-dataset combinations
we tested. Its price was tokens (10 to 925 times more output, model- and domain-dependent) and
reliability: repeated grades of the same answer agreed with each other less in nearly every cell.
Where a grading model collapsed, assigning zero to roughly half of the code answers, reasoning
repaired it; where the model already discriminated, the gain was unpredictable. Evaluation guidance
helped wherever the human reference was trustworthy. That qualifier became a finding in itself: on
the deployment dataset, an unvalidated human reference masked a real rubric benefit, which only
became visible on the stratum of grades with evidence of teacher review. Transfer of the public-dataset results to the deployment
context was partial: the top configuration carried over, but the ranking below it reshuffled. Together
these results yield the central design rule of the framework: configuration priors are starting
points, and a deployment must validate them locally against a reference it trusts.

The contribution is fourfold. First, a configuration decision guide for LLM grading in CS, keyed by
observable task and model behaviour rather than by model identity. Second, the first
grading-specific, within-family isolation of the reasoning effect, measured jointly on agreement,
consistency, and cost. Third, an evidence-backed account of how unreliable reference grades distort
configuration conclusions, with a curation strategy that recovers a usable signal. Fourth, a released
harness, prompt templates, and configuration matrices that make the protocol reproducible. PT-CS
originates from a grading platform built by the first author; we declare this conflict of interest
and the safeguards it requires in Section~\ref{sec:threats}.

The paper is organised as follows. Section~\ref{sec:related} positions the study in the
learning-analytics and LLM-grading literature. Section~\ref{sec:method} describes the design space,
datasets, grading harness, and evaluation protocol. Section~\ref{sec:results} reports the results by
research question. Section~\ref{sec:discussion} synthesises the configuration framework.
Section~\ref{sec:threats} states the threats to validity, and Section~\ref{sec:conclusion} concludes.
