%% sections/threats.tex -- Phase 7.1. Register: CLAUDE.md S9.2. REWRITTEN per PHASES 7.1: the old
%% "two-teacher consensus, human-validated" wording is the S2.3 over-claim and must not reappear.
%% Groups: (A) gold/data (COI, reliability, criterion limit, inflation-as-documented-not-supported),
%% (B) measurement/reproducibility, (C) scope of the claims. Inflation proxy numbers regenerate via
%% phase5.signed_deviation on the declared baseline cell.

\section{Threats to Validity}
\label{sec:threats}

\textbf{Gold and data.} The first author built and owns the platform that produced PT-CS and will
build its successor; this conflict of interest is declared here, and the safeguards below exist
because of it. PT-CS grades are of mixed provenance and limited, uneven reliability: the source
course was a lower-rigour requalification course, not all grades were teacher-reviewed, some may be
unreviewed AI suggestions, and some final grades were adjusted incoherently with their per-criterion
scores. We mitigate by stratification (Section~\ref{sec:datasets}): PT-CS-verified is primary for
transfer and absolute levels, the full set appears only in paired contrasts and in the sensitivity
analysis. The curation criterion has a declared limit: a final grade differing from the criterion sum
proves intervention, not correctness, and an exact-sum grade may still have been reviewed; the
stratum is therefore a conservative proxy, and every transfer claim states its reduced N. Where
teachers intervened, the grade is a consensus correction of an AI-seeded suggestion; no per-teacher
scores exist, so no inter-rater agreement or human ceiling is recoverable. A second, distinct
concern is grade inflation: a lenient course standard would bias the gold even where review is
proven, and the verified stratum does not resolve it. The measurable proxy does not support it: the
signed model-minus-gold deviation in the baseline cell is mixed, lenient on short answers (SemEval
$+0.05$, verified PT-CS theory $+0.10$) and strict on code (RIAYN $-0.19$, Mohler $-0.15$). The
one suggestive signal, verified PT-CS code being strictest ($-0.27$), tracks the zero-collapse, not
the course: the extra strictness relative to RIAYN shrinks as the collapse disappears across models
($-0.077$ Qwen at 49\% zeros; $-0.039$ DeepSeek and $-0.018$ GLM at 13\%). We document the threat
without manufacturing evidence for it; the robust sub-claim is that the unvalidated gold understates
the deviation magnitude (Table~\ref{tab:goldsens}).

\textbf{Measurement and reproducibility.} All numbers are conditional on the served backends:
provider, served model identifier, sampling parameters, and run dates are logged per run, and the
same model name on another provider can behave differently. The transferable claim is the
configuration guidance, not the absolute scores. Temperature 0 is not determinism: 10 to 43\% of
items still vary across repetitions depending on the cell, and the reasoning-related part of that
variability is only partly explained by output length (Section~\ref{sec:discussion}); we do not
present it as purely mechanical. Reasoning cost is not comparable across backends, because DeepInfra
folds reasoning tokens into completion tokens while GPT-5.1 itemises them; premiums are therefore
within-backend ratios. The truncation exclusion is non-random: the 337 excluded run rows (222 items)
are the longest-reasoning Qwen-on cases, so the hardest items drop out and the reasoning benefit is
established on tractable items only. Finally, the bootstrap intervals are reported without
multiple-comparison correction, which is why we enumerate significant cells rather than make
family-wise claims.

\textbf{Scope of the claims.} GPT-5.1 is a reference point, not an inference target: it graded
60-item subsets in the main contrasts and 11 verified short-answer items in transfer, so it
corroborates and is never ranked. Code evidence is narrower than short-answer evidence (two
Java-centric datasets), so code conclusions are stated more tentatively throughout. The guidance
holds across the model families tested, with reasoning isolated within family; it is not a universal
claim about LLMs. The guidance contrast bundles two related interventions (a rubric for the code
datasets, a reference answer for the short-answer ones) and is never pooled across them. The
criterion-decomposition contrast ran with reasoning off only, so the reasoning by decomposition
interaction is unmeasured.

\textbf{Data and code availability.} We release the grading harness, the prompt templates,
the configuration matrices, and the ingest code for the three public datasets, together with the
per-run logs (provider, served model identifier, sampling parameters, timestamps). With this bundle the
public-dataset results (Mohler, SemEval, RIAYN), which carry every cross-dataset and transfer
comparison, reproduce end to end. PT-CS is not shared, in any form, including any anonymised subset,
because it consists of real student answers whose secondary use is bound by institutional ethics and
data-protection rules; we state this as a settled position, not as a release pending request. The
PT-CS-verified curation is instead specified completely in the paper, so that the method, not the data,
is open to scrutiny: the curation criterion is a final grade that differs from the sum of its
per-criterion scores (Section~\ref{sec:datasets}), and the three strata are intervened 32.3\%,
exact-sum 33.2\%, and no-criteria 34.5\% (Table~\ref{tab:strata}). A site holding equivalent
institutional data can therefore reconstruct the stratification, and the full-versus-verified gold
sensitivity, from this specification using the released code.
