%% sota_draft.tex -- Related Work / state-of-the-art draft (Phase 1.6).
%% Framed for IEEE TLT (learning analytics / EDM / assessment-tool register).
%% HARD CAP ~2 pages (CLAUDE.md S9.1): written before results exist; do not let it balloon.
%% \input this into main.tex (Phase 7). Citation keys resolve against article/references.bib.
%% ToE reframing notes are in % comments marked [ToE].

\section{Related Work}
\label{sec:related}

Automated grading is a long-standing concern of the learning-analytics and educational-data-mining
communities, where it is valued less as a stand-alone scoring engine than as an assessment \emph{tool}
that must be reliable, transparent, and fair before it can scale formative or summative feedback. A prior
systematic review of the field~\cite{romano_slr_2025} synthesises this trajectory across five eras --- from
feature-engineered essay scoring, through transformer encoders, to the current generation of large language
models (LLMs) --- and draws one organising conclusion: grading quality is a \emph{task--metric--deployment
triad}, not a single leaderboard, so a strong number on one axis (agreement, error, consistency, or
operational cost) does not imply reliability on another. Everything established up to mid-2025 is taken from
that review; here we focus on what the most recent evidence implies for \emph{which configuration} a grader
should use, the question this paper addresses.

\textbf{LLMs as graders: capable but inconsistent.}
Prompted LLMs now match or approach human raters on well-specified short-answer and essay tasks, yet their
reliability is the recurring caveat. Across rubric-conditioned settings, agreement is strong but
\emph{model-} and \emph{prompt-dependent}: consistency varies sharply between systems (e.g.\ open-source
models span ICC $0.29$--$0.83$ on the same task~\cite{moazzez_assessing_2024}), repeated scoring drifts over
time~\cite{pack_large_2024}, and minimally-prompted grading can stay within one rubric point of teachers yet
yield low summative agreement~\cite{tate_can_2024}. Two design levers recur as decisive. First, evaluation
\emph{guidance}: clearer, more explicit rubrics raise consistency~\cite{garcia-huertes_leveraging_2024}, and
closed models with rubric grounding outperform open ones while remaining more lenient than human
raters~\cite{sessler_canai_2025}. Second, \emph{repetition}: because single-shot scores are unstable,
temporal-drift evidence~\cite{pack_large_2024} motivates reporting variance across runs rather than a point
estimate. Reliability is not optional here: grading errors can themselves harm learning, as false positives
and negatives distort how students revise their work~\cite{li_am_2023}. These two observations directly
motivate our evaluation-guidance (RQ2) and consistency (RQ5) axes.

\textbf{Does reasoning help grading? An open question.}
The headline uncertainty concerns \emph{extended reasoning}. In the general LLM-as-a-judge setting, explicit
reasoning improves judgement: using a within-family thinking toggle on Qwen3, ``thinking'' models gain
roughly ten accuracy points at modest extra cost~\cite{jayarao_explicit_2025}. The benefit, however, is
\emph{task- and architecture-dependent} --- on adjacent natural-language-generation evaluation (machine
translation and summarisation), one reasoning model improves with extended thinking while another
\emph{underperforms} its own non-reasoning variant~\cite{larionov_deepseek_2025} --- and the
``overthinking'' literature shows that long reasoning traces add redundant computation and can degrade
calibration-sensitive tasks~\cite{sui_stop_2025}. Grading a student answer is exactly such a task: it is
bounded by a rubric and sensitive to score calibration rather than to open-ended deliberation. Whether
reasoning helps or hurts here is therefore \emph{a priori} uncertain. Within educational short-answer grading
itself, the task is increasingly framed as a move ``from memorization to reasoning''~\cite{cong_memorization_2026},
yet that work studies fine-tuning and rubric-based prompting rather than isolating the effect of \emph{model}
reasoning, and prior comparisons confound reasoning with model identity by pitting reasoning and non-reasoning
systems from different vendors. This paper resolves the question empirically with a within-family on/off toggle
(RQ1), the central contribution motivating the study.

\textbf{Code grading with rubrics.}
Our code domain is narrower and less mature than short-answer grading. Locally-run open-source LLMs are
already used to \emph{assist} grading in introductory computer science, positioned as teaching-assistant
support rather than as final graders~\cite{mazzone_exploring_2024}, and reviews of programming education
frame the task as dynamic (test-case) versus static (code-quality) grading with semi-automatic
triage~\cite{manorat_artificial_2025}. Rubric-based code graders that award partial credit reach about
85\% micro-accuracy but still leave roughly a fifth of cases mis-scored --- too high for full
automation~\cite{havare_ai-based_2025}. Most relevant to our framing, question-specific rubrics
substantially outperform question-agnostic or no-rubric prompting on object-oriented and data-structures
code (e.g.\ Cohen's $\kappa$ rising from $\approx0.41$ with no rubric to $0.60$--$0.65$ with a
question-specific one)~\cite{pathak_rubric_2025}, tying the evaluation-guidance dimension directly to the
code setting. Because public code datasets with \emph{human} rubric grades are scarce, our code-domain
conclusions are stated more tentatively than the short-answer ones.

\textbf{Datasets and published baselines as context.}
We develop on established public short-answer corpora --- the Mohler dataset of graded computer-science
answers~\cite{mohler_text_2009,mohler_learning_2011} and the SemEval-2013 Task~7 SciEntsBank/Beetle
sets~\cite{dzikovska_semeval_2013} --- and confirm transfer on a Portuguese deployment context.
Table~\ref{tab:published-baselines} collects reported per-dataset performance, from supervised transformers
(e.g.\ ASAP-SAS QWK $\approx0.79$~\cite{kumar_get_2019}; recent essay QWK $0.81$--$0.83$~\cite{dadi_robust_2025,elmassry_systematic_2025})
to prompted and fine-tuned LLMs on our own sets~\cite{aminah_evaluating_2025,chen_cemft_2026,dada_iattention_2025,fischer_evaluation_2025,duong_automatic_2024,ferreira_mello_automatic_2025,kazi_enhancing_2023,henkel_roars_2025}.
Crucially, these numbers were produced under \emph{different models, prompts, and protocols}; they situate
our study but are \emph{not} its success criterion, which comes from our own like-for-like comparisons. The
literature also illustrates why protocol matters: a reported QWK of $0.95$ on Mohler is an artefact of
silently dropping unparseable answers~\cite{aminah_evaluating_2025}, underscoring our use of an
extractable-rate ($\pi$) metric and our refusal to coerce unparseable output to zero. Dedicated fine-grained
benchmarks for LLM short-answer scoring are also beginning to appear~\cite{lai_sasbench_2026}, but none yet
isolates the configuration axes we study.

\textbf{From model rankings to a configuration framework.}
A growing body of work compares LLMs \emph{as graders} --- large side-by-side studies of programming
submissions~\cite{jukiewicz_comparison_2026}, deployed course evaluations~\cite{policar_automated_2025}, and
multi-model frameworks emphasising transparency and reproducibility~\cite{anghel_graderassist_2025}.
This paper is deliberately \emph{not} another such ranking. Following the design-insight orientation that
TLT asks of evaluation-centred work, our contribution is a \emph{configuration framework}: an empirically
grounded decision guide mapping task characteristics (domain, rubric availability, budget) to a recommended
grader configuration and its cost. This complements the learning-analytics literature on responsible,
auditable LLM grading --- structured ``grading events'' with local hosting and fine-tuning on course
data~\cite{arz_von_straussenburg_enabling_2025} and embedding-based short-answer
grading~\cite{putnikovic_embeddings_2023} --- and feeds the design decisions of a deployed assessment
platform. Throughout, we treat the human reference as a teacher-validated consensus grade (not independent
double annotation) and declare the conflict of interest arising from the platform context, in line with the
ground-truth-honesty requirements such deployments demand.

% [ToE] If the venue becomes IEEE Transactions on Education, re-foreground the pedagogical contribution:
% lead with instructor workload, assessment validity, fairness, and the computing-education context;
% expand the human-in-the-loop and false-positive-harm argument (cf. \cite{li_am_2023}); keep the
% model/benchmark detail terser. The study itself is unchanged.
